Probabilities and propensities: conversion experiences in online grocery shopping

### 2017-07-17 lists_conversion_experience.rmd
starting reconstruction

While Tesco  succeeded in data-mining its customers using demographic segmentation,  and perhaps became the UK's biggest supermarket with the help of data-mining in the 1990s, the shopping environment  in 2017 is markedly different [@Turow_2017]. It is no longer organised around campaigns involving special offers or redemption of points for demographically segmented loyal customers (DunnHumby made heavy use of UK Census data). It can no longer rely on placement of goods in carefully chosen locations in stores. It needs or at least might want to continuously up-sell and cross-sell to customers who might hardly ever visits the supermarket itself.

How could we characterise the shift from ClubCard data mining to online grocery markets?  If 'Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet' [How SEO helps Tesco to dominate the online grocery market](https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market)[@Silverwood-Cope_2014], then Tesco itself has undergone some kind of conversion? Patel described her work at DunnHumby as shifting the recommender system from a 'rules-based list' to a 'relevancy model.' The relevancy model changed how the 'Have You Forgotten' list was constructed.   This shift is a typical of broader re-organisation of prediction in which existing data analysis practices are being re-distributed and intensified in particular ways. The shift in models results in a more probabilistic structuring of lists.  The ostensible banality of the 'Have you forgotten?' recommendations betrays micrological signs of a topologically complex predictive infrastructure that constructs alignments between people and things in significant ways.

>>>>

While Tesco  succeeded in data-mining its customers using demographic segmentation,  and perhaps became the UK's biggest supermarket with the help of data-mining in the 1990s, the shopping environment  in 2017 is markedly different [@Turow_2017]. It is no longer organised around campaigns involving special offers or redemption of points for demographically segmented loyal customers (DunnHumby made heavy use of UK Census data). It can no longer rely on placement of goods in carefully chosen locations in stores. It needs or at least might want to continuously up-sell and cross-sell to customers who might hardly ever visit the supermarket itself.

How could we characterise the shift from _ClubCard_ data mining to online grocery markets?  If 'Tesco is the clear winner in the online grocery market, in fact it takes almost 50p of every £1 spent on food shopping on the internet' [@Silverwood-Cope_2014](https://econsultancy.com/blog/64841-how-seo-helps-tesco-to-dominate-the-online-grocery-market), then Tesco itself has undergone some kind of conversion? Patel described her work at DunnHumby as shifting the recommender system from a 'rules-based list' to a 'relevancy model.' The relevancy model changed how the 'Have You Forgotten' list was constructed.   This shift is a typical of broader re-organisation of prediction in which existing data analysis practices are being re-distributed and intensified in particular ways. The shift in models results in a more probabilistic structuring of lists.  The ostensible banality of the 'Have you forgotten?' recommendations betrays micrological signs of a topologically complex predictive infrastructure that constructs alignments between people and things in significant ways.

### 2017-07-17 log.rmd
starting reconstruction

>>>>

## Thu 13 Jul 2017 15:29:02 BST
- starting revisions by going through ref2; 
- philosophico-methodological preramble: TODO: put this as opening paragraph

> This paper is a philosophical reconstruction [@Dewey_2017] of an online grocery ordering system. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how recommendations were created for online grocery orders. Given this highly mediated and limited encounter with a recommender system and one of its designers, the paper constructs a multi-layered account.   It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping. It conducts several code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some important elements of the system in question.  It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map the functional statements and practices underpinning the knowledge and order generated by such systems over time. Not least, I draw on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations around probabilities and their enactments. Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Rabinow_2004; @Bogost_2012], through this combination of approaches,  the paper itself provisionally allows its own reconstruction to be shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophical engagement with its objects may well seem overly concerned with the technical minutiae of the grocery ordering system, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the main philosophical argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a more pervasive role.

## Fri 14 Jul 2017 11:30:07 BST
- read popper on propensities again, updated notes in ~/notes/popper_propensities.md, and also marked up the paper copy more;  
- TODO: for methods add sandvig on audit [@Sandvig_2015a]

> The main argument of the paper concerns how  probability operates in Big Data situations. While calculations of probability have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on a much thicker situational texture. Operational probabilities differ from epistemic probabilities in their position in the world, as part of infrastructures, transactions, and event-structures. Taking their complex siting into account, and accepting their _physical_ reality, we might come to understand the world they belong to as what the philosopher of science Karl Popper understood as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. Given this understanding of operational probability, our understanding of what is enacted in contemporary data analytics, especially in association with platforms [@Gillespie_2010] might shift.

> The broad implication of the argument concerns how we think about social order or regularities: proponents and critics of Big Data present the advent of large volumes of data and analytical techniques (machine learning in particular) as transforming organisations, institutions, economic processes and social life. I suggest too that a transformation or conversion event is occurring, but one which might open up the possibility of experiencing the world differently, and conceiving of social ordering processes somewhat afresh.  Even if the argument is centred on an eminently social practice such as shopping, the analysis of how a world of propensities continuously and unevenly converts into realizations applies across the board, from the most physical to the most ideal or abstract settings. The average everydayness of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- particle physics for instance -- do not.

## Mon 17 Jul 2017 11:46:32 BST
- still not sure how radically to restructure this. 
- the simplest order I can think of
    - A. Introduction - includes a,b,c from above
    - B. Patel section + opening description
    - C. Shopping and lists -- Miller, and list section
    - D. Archaeology of recommendations -- from 1984 -2007
    - E. Experimenting with probabilistic conversions
    - F. Conclusion
- will see if this structure works with a rough cut and paste
- re-ordering has gone ok to the end of D. But section E is really tricky -- this is the conversion event; do I include the machine learning stuff as part of D; or use it to be begin E?
    - E: shift from apriori to logistic regression; then think of good features; experimentalising/trials; infrastructural operationalisation; but then what about substitution; complementarity? All of this section is meant to be about probabilisation? And the propensities idea;    
- TODO: write short paragraph describing my overall social theory of lists (approx 200 words)
- new title: Probabilities in the data analytic conversion of social order
- TODO: play with the structuring of Section E. 
-
